Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-05-08 linear_iso.tex
further refinements to appendix D2, D3. plus minor notation update
mkl going through entire manucript, merge with youdin's updates

\item We make the low frequency approximation, assuming $|\upsilon^2|\ll \kappa^2,\,\Omega^2$. 
%keeping only the lowest order terms in $|\upsilon|/\kappa$.  
Similar to the incompressible \citepalias{goldreich67} or anelastic \citepalias{nelson13,barker15} approximations,
the low frequency approximation filters p-modes in favor of g-modes, i.e.\ inertial
gravity waves \citep{lubow93}.

@@

 In Appendix \ref{global_corr}  we discuss the limits of these approximations
 and show that they are valid in the parameter space of interest for the VSI.
In particular we point out that the fully-radially-local approximation is only valid for
short cooling times ($\beta \lesssim 1$).

>>>>

\item We make the low frequency approximation, assuming $|\upsilon^2|\ll \kappa^2,\,\Omega^2$. 
%keeping only the lowest order terms in $|\upsilon|/\kappa$.  
Similar to the incompressible \citepalias{goldreich67} or anelastic \citepalias{nelson13,barker15} approximations,
the low frequency approximation filters acoustic waves in favor of
inertial-gravity waves \citep{lubow93}.

@@

 In Appendix \ref{global_corr}  we discuss the limits of these approximations
 and show that they are valid in the parameter space of interest for the VSI.
In particular we point out that the fully-radially-local approximation is only valid for
short cooling times, $\beta \lesssim O(1)$.

### 2015-05-08 numerical.tex
mkl going through entire manucript, merge with youdin's updates

In Fig. \ref{compare_modes_vnoniso_kx10} we plot the mode diagram for
$\khat=30$ for several values of $\beta$. This plot is 
qualitatively similar to Fig. \ref{compare_modes_cool_kx10}. As before
we find as $\beta$ is increased, higher order body modes are rapidly
stabilized, eventually only the fundamental mode can operate before
the VSI is entirely suppressed. (Note the significant reduction in
growth rates for $\beta=0.4$, which is close to
$\beta_\mathrm{crit,noniso}$.) However, this behavior is not seen for 
large wavenumbers (see below).

>>>>

In Fig. \ref{compare_modes_vnoniso_kx10} we plot the mode diagram for
$\khat=30$ for several values of $\beta$. This plot is 
qualitatively similar to Fig. \ref{compare_modes_cool_kx10}. As before
we find as $\beta$ is increased, higher order modes are rapidly
stabilized, eventually only the fundamental mode can operate before
the VSI is entirely suppressed. (Note the significant reduction in
growth rates for $\beta=0.4$, which is close to
$\beta_\mathrm{crit,noniso}$.) However, this behavior is not seen for 
large wavenumbers (see below).

### 2015-05-08 paper.tex
mkl going through entire manucript, merge with youdin's updates

\title{Cooling Requirements for the Vertical Shear Instability 
  in Protoplanetary Disks}
%\title{Vertical shear instability in astrophysical disks: vertically
%global linear calculations with thermal relaxation and applications
%to the Solar Nebula}  
\author{Min-Kai Lin \& Andrew N Youdin}
\affil{Department of Astronomy and Steward Observatory, \\
University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\email{minkailin@email.arizona.edu, youdin@email.arizona.edu}

@@

For vertically global disturbances considered in this work, 
% To roughly obtain a global cooling time criterion, we evaluate Eq.\ \ref{bcrit_est}
% at $z \sim H$.  To get the precise form we want, we choose $z = \gamma H / 2$ to 
we obtain the cooling requirement 
\begin{align}\label{prelim_bcrit}
  t_c <  h |q| \OmK^{-1}/ (\gamma - 1)
\end{align}   
for the VSI to operate effectively in a vertically isothermal, thin
disk. We derive this result more formally in
\S\ref{iso_vsi_beta_crit}, but notice that Eq. \ref{prelim_bcrit}
may be obtained from Eq. \ref{bcrit_est} by evaulating it at the
height $|z|=\gamma H/2$. %i.e., at $z=O(H)$

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.7cm 0cm
  0.73cm]{figures/bcrit_mmsn2d_kx10}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.46cm 0cm
  0.73cm]{figures/bcrit_mmsn2d_kx50}
  \caption{Estimate of the thermal timescale  in a fiducial 
    PPD model, the Minimum Mass Solar Nebula. The red
    line corresponds to a characteristic thermal timescale, derived in
    this work, below which the VSI is expected to operate effectively
    (the region outside the line).    
    \label{bcrit_mmsn2d} 
  }
\end{figure}

We now consider whether, and where, thermal timescales in realistic PPDs
are short enough to support VSI growth.  Fig. \ref{bcrit_mmsn2d} plots the
 thermal timescales in our fiducial PPD model with a small grain dust opacity. 
 Since cooling by radiative diffusion depends on a perturbation's lengthscale, we
 plot $t_c$ for two values of the radial wavenumber,  $k_x$,  the relevant 
 component for VSI modes that are radially narrow.  The red line
marks the boundary implied by Eq. \ref{prelim_bcrit}.  We caution that 
 Eq. \ref{prelim_bcrit} applies when $t_c$ is vertically uniform.  This section is 
 aimed at basic insights; while more realistic spatial variations in
 $t_c$ is considered in \S\ref{application}. 
 %analyzes the VSI with realistic spatial variations in.

 Due to the smaller values of $t_c$ in the outer disk, we generally expect the VSI to 
 operate outside $\sim 10$ AU.  This conclusion changes however, if
 the dust opacity is reduced,  
slowing down cooling, as we show in \S\ref{application}. % (e.g.\ Fig.\
% \ref{mmsn_bcrit_bcool}). 
In the inner disk, thermal times are long in the midplane, but short at higher altitutdes. 
Again, this structure necessitates the more detailed analyses that follow.  Nevertheless, we can already 
see that due to radiative diffusion higher wavenumbers can cool more rapidly.  Thus smaller scale disturbances 
may trigger VSI at smaller disk radii.

On the other hand, one cannot obtain arbitrarily short cooling times
by invoking arbitrarily small scales (cf. \citealt{urpin03}, 
\citetalias{barker15}), because very small wavelengths are in the
optically thin cooling regime,  where cooling is independent of
lengthscale. Furthermore, 
%while we do not include viscosity in our
%idealized disk models, in reality 
small-scale perturbations will be 
prone to viscous damping, including the effective viscosity from turbulent motions.
While our linear, inviscid calculations do not address this issue directly, we expect that 
shorter wavelength modes will drive weaker non-linear (and non-axisymmetric) transport.

>>>>

\title{Cooling Requirements for the Vertical Shear Instability 
  in Protoplanetary Disks}
%\title{Vertical shear instability in astrophysical disks: vertically
%global linear calculations with thermal relaxation and applications
%to the Solar Nebula}  
\author{Min-Kai Lin\altaffilmark{$\dagger$} \& Andrew N Youdin}
\affil{Department of Astronomy and Steward Observatory, \\
University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\altaffiltext{$\dagger$}{Steward Theory Fellow}
\email{minkailin@email.arizona.edu, youdin@email.arizona.edu}

@@

For vertically global disturbances considered in this work, 
% To roughly obtain a global cooling time criterion, we evaluate Eq.\ \ref{bcrit_est}
% at $z \sim H$.  To get the precise form we want, we choose $z = \gamma H / 2$ to 
we obtain the cooling requirement 
\begin{align}\label{prelim_bcrit}
  t_c <  &\underbrace{\frac{h |q| }{\gamma - 1}}_{\beta_\mathrm{crit}}\OmK^{-1}
 % &\equiv \beta_\mathrm{crit}\OmK^{-1},\notag
\end{align}   
for the VSI to operate effectively in a vertically isothermal, thin
disk. We derive this result more formally in
\S\ref{iso_vsi_beta_crit}, but notice that Eq. \ref{prelim_bcrit}
may be obtained from Eq. \ref{bcrit_est} by evaulating it at the
height $|z|=\gamma H/2$. %i.e., at $z=O(H)$

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.7cm 0cm
  0.73cm]{figures/bcrit_mmsn2d_kx10}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.46cm 0cm
  0.73cm]{figures/bcrit_mmsn2d_kx50}
  \caption{Estimate of the thermal timescale  in a fiducial 
    PPD model, the Minimum Mass Solar Nebula. The red
    lines corresponds to a maximum thermal timescale, derived in
    this work, below which the VSI is expected to operate effectively
    (the region bounded by the red lines).     
    \label{bcrit_mmsn2d} 
  }
\end{figure}

We now consider whether, and where, thermal timescales in realistic PPDs
are short enough to support VSI growth.  Fig. \ref{bcrit_mmsn2d} plots the
thermal timescales in our fiducial PPD model with a small grain dust opacity. 
Since cooling by radiative diffusion depends on a perturbation's lengthscale, we
plot $t_c$ for two values of the radial wavenumber,  $k_x$,  the relevant 
component for VSI modes that are radially narrow. In the
optically-thin cooling regime, $t_c$ is independent of lengthscales. 
The red lines mark the boundaries implied by
Eq. \ref{prelim_bcrit}.  We caution that 
Eq. \ref{prelim_bcrit} applies when $t_c$ is vertically uniform.  This section is 
aimed at basic insights; while more realistic spatial variations in
$t_c$ is considered in \S\ref{application}.

Due to the smaller values of $t_c$ in the outer disk, we generally expect the VSI to 
operate outside $\sim 10$AU, but inside $\sim 150$AU since $t_c$
exceeds the critical value beyond this radius. This conclusion changes however, if
the dust opacity is reduced,  
slowing down cooling, as we show in \S\ref{application}. % (e.g.\ Fig.\
% \ref{mmsn_bcrit_bcool}). 
In the inner disk, thermal times are long in the midplane, but short at higher altitutdes. 
Again, this structure necessitates the more detailed analyses that follow.  Nevertheless, we can already 
see that due to radiative diffusion higher wavenumbers can cool more rapidly.  Thus smaller scale disturbances 
may trigger VSI at smaller disk radii.

On the other hand, one cannot obtain arbitrarily short cooling times
by invoking arbitrarily small scales (cf. \citealt{urpin03}, 
\citetalias{barker15}), because very small wavelengths are in the
optically thin cooling regime and there exists a minimum thermal
timescale. Furthermore, 
%while we do not include viscosity in our
%idealized disk models, in reality 
small-scale perturbations will be 
prone to viscous damping, including the effective viscosity from turbulent motions.
While our linear, inviscid calculations do not address this issue directly, we expect that 
shorter wavelength modes will drive weaker non-linear (and non-axisymmetric) transport.

### 2015-05-08 setup.tex
mkl going through entire manucript, merge with youdin's updates

\subsection{Thermal relaxation}
We model radiative effects as thermal
relaxation with a timescale $t_c$, i.e.,
\begin{align}\label{thermal_relax}
  \Lambda  \equiv \frac{\rho\mathcal{R}}{\mu}\frac{T - T_0}{t_c},
\end{align}
where $T_0=T_0(r,z)$ is the equilibrium temperature.  We 
define the dimensionless cooling time 
\begin{align}\label{beta_def}
  \beta \equiv \OmK t_c
\end{align}
relative to the Keplerian frequency, $\OmK=\sqrt{GM_*/r^3}$.  We 
use the terms `thermal relaxation' and `cooling' interchangeably
throughout this paper.

@@

\subsection{Baroclinic disk equilibria}\label{eqm}
The equilibrium disk is steady and axisymmetric with density,
pressure and rotation profiles $\rho(r,z)$, $P(r,z)$ and
$\Omega(r,z)$, respectively. These functions satisfy  
\begin{subequations}\begin{align}
  0 &= \frac{1}{\rho}\frac{\p P}{\p z} + \frac{\p \Phi_*}{\p z},\label{vert_equilibrium}\\
  r\Omega^2&= \frac{1}{\rho}\frac{\p P}{\p r} + \frac{\p \Phi_*}{\p
    r}.\label{hori_equilibrium} 
\end{align} \end{subequations}
A unique solution requires additional assumptions about disk structure.  We choose
\begin{subequations}\begin{align}
 \rho(r,0) &\equiv \rho_0(r) = \rho_{00}\left(\frac{r}{r_0}\right)^p, \\
 T(r,0) &\equiv T_0(r) = T_{00}\left(\frac{r}{r_0}\right)^q, \\
 P(r,z) & = K(r)\rho^{\Gamma}(r,z). \label{eqm_press}
\end{align} \end{subequations}
where $r_0$ is a fiducial radius and $p$ and $q$ are the standard powerlaw indices
for midplane density and temperature, respectively.  The polytropic index $\Gamma$ 
parametrizes the vertical stratification with $\Gamma = 1$ ($\Gamma = \gamma$)
describing vertically isothermal (adiabatic) disks, respectively.  The ideal gas
law requires $K = \rho_0^{1-\Gamma}\mathcal{R} T_0/\mu$.

@@

The vertically isothermal case, $\Gamma = 1$, can be calculated either 
as a special case or as the $\Gamma \rightarrow 1$ limit of Eq. \ref{eqm_dens},    
\begin{subequations}\begin{align}
\lim_{\Gamma\to1}  \frac{\rho(r,z)}{\rho_0(r)} &=
  \exp{\left[\frac{1}{ h^2}\left(\frac{1}{\sqrt{1+z^2/r^2}}-1\right)\right]},   \\
  & \simeq \exp {\left( - {z^2 \over 2 H^2}\right)} \quad\text{for } |z|\ll r \label{rhoisothin}. 
  \end{align}\end{subequations}
Eq.\ \ref{rhoisothin} is the approximate form of the density field in the thin-disk limit.  
We will primarily focus on 
vertically isothermal disks, the relevant case for passively irradiated
PPDs \citep{chiang97}.

@@

A more general expression for vertical shear holds for vertical polytopes 
(which satisfy Eq.\ \ref{eqm_press}), with no assumed radial structure:
\begin{align}
  r\frac{\p\Omega^2}{\p z} = - \rho^{\Gamma-1}\frac{\p\ln\rho}{\p
    z}\frac{dK}{dr} = - \frac{g_z}{\Gamma}\frac{d\ln K}{dr}, \label{vertical_shear}
\end{align}
where $g_z = \p_z P/\rho$.

@@

\subsection{Entropy gradients and vertical buoyancy}
The gradients of entropy, $S\equiv
C_P\ln{P^{1/\gamma}/\rho}$, in our disk models are
\begin{subequations}
\begin{align}
  &\frac{\p S}{\p r} = C_P\left[\frac{s}{\gamma r} +
    \left(\frac{\Gamma}{\gamma}-1\right)\frac{\p \ln{\rho}}{\p
       r}\right], \label{dSdr}\\
  &\frac{\p S}{\p z} = C_P
  \left(\frac{\Gamma}{\gamma}-1\right)\frac{\p \ln{\rho}}{\p z} = {C_P g_z \over c_s^2} {\Gamma - \gamma \over \gamma},\label{dSdz}
\end{align} 
\end{subequations}
 where $C_P$ is the heat capacity at
constant pressure.

@@

\subsection{Stability without cooling}\label{solberg}
In the absence of cooling, axisymmetric stability is ensured
if both Solberg-Hoiland criteria  are satisfied:
\begin{align}
  \kappa^2 - \frac{1}{\rho C_P}\nabla P \cdot \nabla S &> 0,\\
  -\frac{\p P}{\p z}\left(\frac{\p j^2}{\p r}\frac{\p S}{ \p z} -
    \frac{\p j^2}{\p z}\frac{\p S}{\p r} \right)&>0. \label{second_SH} 
\end{align}
\citep{tassoul78}. %, where $j=r^2\Omega$ is the specific angular
%momentum. %and $\kappa = r^{-3} \p_r j^2 $ is the epicyclic frequency.  
For rotationally-supported, thin disks %  pressure gradients are small 
% compared to rotation, so 
the first criterion is generally satisfied. 
Thus, to assess stability, we consider the second
criterion. 
Without loss of generality, we evaluate Eq. \ref{second_SH} in $z>0$
where $\p_zP<0$. Here we approximate $\p_r j^2\simeq r^3\OmK^2$ and
use Eq. \ref{vertical_shear} for $\p_zj^2$. Then 
\begin{align}\label{solberg2}
  % \gamma - \Gamma > \frac{s h^2}{\Gamma}
  % \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s
  %   -\left(\gamma-\Gamma\right)\frac{\p\ln{\rho}}{\p\ln{r}} \right]. 
  \gamma - \Gamma > \frac{ h^2}{\Gamma}
  \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s^2
    -s\left(\gamma-\Gamma\right)\frac{\p\ln{\rho}}{\p\ln{r}} \right]
\end{align} 
implies stability. For typical model parameters the right hand
side (RHS) is $O( h^2)$. Hence, instability in the adiabatic limit  
requires $\gamma < \Gamma + O( h^2)$. Since we consider
$ h\ll1$, this implies the equilibrium disk must have a small
entropy gradient in order to be unstable to axisymmetric adiabatic
perturbations. This is generally not true for passively irradiated, stably
stratified disks \citep{chiang97}, in which the VSI requires rapid
cooling, as shown below.

@@

\subsection{Radially local approximation}\label{sec:radlocal}
We consider axisymmetric perturbations to the above disk equilibria.    
The growth of the VSI is strongest for short radial wavelengths, as compared
 to the disk radius \citepalias{nelson13,barker15}.  We thus perform a
 standard two step process to obtain linearized equations in the radially local 
 approximation (e.g., \citealp{goldreich67}, who also consider vertically local perturbations).

>>>>

\subsection{Thermal relaxation}
We model radiative effects as thermal
relaxation with a timescale $t_c$, i.e.,
\begin{align}\label{thermal_relax}
  \Lambda  \equiv \frac{\rho\mathcal{R}}{\mu}\frac{T - T_\mathrm{eq}}{t_c},
\end{align}
where $T_\mathrm{eq}=T_\mathrm{eq}(r,z)$ is the equilibrium temperature.  We 
define the dimensionless cooling time 
\begin{align}\label{beta_def}
  \beta \equiv \OmK t_c
\end{align}
relative to the Keplerian frequency, $\OmK=\sqrt{GM_*/r^3}$.  We 
use the terms `thermal relaxation' and `cooling' interchangeably
throughout this paper.

@@

\subsection{Baroclinic disk equilibria}\label{eqm}
The equilibrium disk is steady and axisymmetric with density,
pressure and rotation profiles $\rho(r,z)$, $P(r,z)$ and
$\Omega(r,z)$, respectively. These functions satisfy  
\begin{subequations}\begin{align}
  0 &= \frac{1}{\rho}\frac{\p P}{\p z} + \frac{\p \Phi_*}{\p z},\label{vert_equilibrium}\\
  r\Omega^2&= \frac{1}{\rho}\frac{\p P}{\p r} + \frac{\p \Phi_*}{\p
    r}.\label{hori_equilibrium} 
\end{align} \end{subequations}
A unique solution requires additional assumptions about disk structure.  We choose
\begin{subequations}\begin{align}
 \rho(r,0) &\equiv \rho_0(r) = \rho_{00}\left(\frac{r}{r_0}\right)^p, \\
 T(r,0) &\equiv T_0(r) = T_{00}\left(\frac{r}{r_0}\right)^q, \\
 P(r,z) & = K(r)\rho^{\Gamma}(r,z). \label{eqm_press}
\end{align} \end{subequations}
where $r_0$ is a fiducial radius and $p$ and $q$ are the standard powerlaw indices
for midplane density and temperature, respectively. We consider $q<0$
without loss of generality.  The polytropic index $\Gamma$ 
parametrizes the vertical stratification with $\Gamma = 1$ ($\Gamma = \gamma$)
describing vertically isothermal (adiabatic) disks, respectively.  The ideal gas
law requires $K = \rho_0^{1-\Gamma}\mathcal{R} T_0/\mu$.

@@

The vertically isothermal case, $\Gamma = 1$, can be calculated either 
as a special case or by taking the limit of Eq. \ref{eqm_dens},    
\begin{subequations}\begin{align}
\lim_{\Gamma\to1}  \frac{\rho(r,z)}{\rho_0(r)} &=
  \exp{\left[\frac{1}{ h^2}\left(\frac{1}{\sqrt{1+z^2/r^2}}-1\right)\right]},   \\
  & \simeq \exp {\left( - {z^2 \over 2 H^2}\right)} \quad\text{for } |z|\ll r \label{rhoisothin}. 
  \end{align}\end{subequations}
Eq.\ \ref{rhoisothin} is the approximate form of the density field in the thin-disk limit.  
We will primarily focus on 
vertically isothermal disks, the relevant case for passively irradiated
PPDs \citep{chiang97}.

@@

A more general expression for vertical shear holds for vertical polytropes 
(which satisfy Eq.\ \ref{eqm_press}), with no assumed radial structure:
\begin{align}
  r\frac{\p\Omega^2}{\p z} = - \rho^{\Gamma-1}\frac{\p\ln\rho}{\p
    z}\frac{dK}{dr} = - \frac{g_z}{\Gamma}\frac{d\ln K}{dr}, \label{vertical_shear}
\end{align}
where $g_z = \p_z P/\rho$.

@@

\subsection{Entropy gradients and vertical buoyancy}
The gradients of specific entropy, $S\equiv
C_P\ln{P^{1/\gamma}/\rho}$, in our disk models are
\begin{subequations}
\begin{align}
  &\frac{\p S}{\p r} = C_P\left[\frac{s}{\gamma r} +
    \left(\frac{\Gamma}{\gamma}-1\right)\frac{\p \ln{\rho}}{\p
       r}\right], \label{dSdr}\\
  &\frac{\p S}{\p z} = C_P
  \left(\frac{\Gamma}{\gamma}-1\right)\frac{\p \ln{\rho}}{\p z} = {C_P g_z \over c_s^2} {\Gamma - \gamma \over \gamma},\label{dSdz}
\end{align} 
\end{subequations}
 where $C_P$ is the heat capacity at
constant pressure.

@@

\subsection{Stability without cooling}\label{solberg}
In the absence of cooling, axisymmetric stability is ensured
if both Solberg-Hoiland criteria  are satisfied:
\begin{align}
  \kappa^2 - \frac{1}{\rho C_P}\nabla P \cdot \nabla S &> 0,\\
  -\frac{\p P}{\p z}\left(\frac{\p j^2}{\p r}\frac{\p S}{ \p z} -
    \frac{\p j^2}{\p z}\frac{\p S}{\p r} \right)&>0. \label{second_SH} 
\end{align}
\citep{tassoul78}. %, where $j=r^2\Omega$ is the specific angular
%momentum. %and $\kappa = r^{-3} \p_r j^2 $ is the epicyclic frequency.  
For rotationally-supported, thin disks %  pressure gradients are small 
% compared to rotation, so 
the first criterion is generally satisfied. 
Thus, to assess stability, we consider the second
criterion. 
Without loss of generality, we evaluate Eq. \ref{second_SH} in $z>0$
where $\p_zP<0$. Here we approximate $\p_r j^2\simeq r^3\OmK^2$ and
use Eq. \ref{vertical_shear} for $\p_zj^2$. Then 
\begin{align}\label{solberg2}
  % \gamma - \Gamma > \frac{s h^2}{\Gamma}
  % \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s
  %   -\left(\gamma-\Gamma\right)\frac{\p\ln{\rho}}{\p\ln{r}} \right]. 
  \gamma - \Gamma > \frac{ h^2}{\Gamma}
  \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s^2
    -s\left(\gamma-\Gamma\right)\frac{\p\ln{\rho}}{\p\ln{r}} \right]
\end{align} 
implies stability. For typical model parameters, 
Eq. \ref{solberg2} indicates that axisymmetric instability in the adiabatic limit  
requires $\gamma < \Gamma + O( h^2)$, corresponding to a small entropy 
gradient.  
%This implies the equilibrium
%disk must have a small 
%entropy gradient in order to be unstable to axisymmetric adiabatic
%perturbations. 
This is generally not true for passively irradiated, stably
stratified disks \citep{chiang97}, in which the VSI requires rapid
cooling, as shown below.

@@

\subsection{Radially local approximation}\label{sec:radlocal}
We consider axisymmetric perturbations to the above disk equilibria.    
The growth of the VSI is strongest for short radial wavelengths, as compared
 to the disk radius \citepalias{nelson13,barker15}.  We thus perform a
 standard two step process to obtain linearized equations in the radially local 
 approximation (e.g., \citealp{goldreich67}, who also consider vertically localized perturbations).

### 2015-05-08 summary.tex
mkl going through entire manucript, merge with youdin's updates

\section{Summary and discussion}\label{summary}
In this paper we study the vertical shear instability (VSI) with a focus the 
role of radiative cooling.  In turn we assess the viability of the VSI 
as an angular momentum transport mechanism in protoplanetary disks (PPDs).

@@

In applying our results to PPDs, we pay particular attention to the transition from 
optically thick radiative diffusion to optically thin Newtonian cooling.  The largest obstacle to VSI occurs in 
high density inner disk, $\lesssim 1$ -- 5 AU, where radiative diffusion times are slow.  Shorter wavelength disturbances
 cool faster, so long as they remain optically thick.  In the inner disk, however, our VSI cooling criterion requires 
 wavelengths that are too short to drive significant transport.  The best hope for the VSI in inner disk appears to 
 be a low surface density, which speeds radiative diffusion by lengthening the photon mean free path.  This option naturally begs 
 the question of what accretion mechanism lowered the surface density in the first place.

 In the outer disk, the VSI tends to cool in the optically thin limit.  The main issue is whether the opacity 
 is high enough for optically thin cooling to be sufficiently fast.  Our standard opacity assumes a Solar abundances of small dust.
 In this case, the VSI can operate from $\sim 5$ -- 100 AU.  A factor of 10 reduction in the opacity, for instance by locking
 small dust into planetesimals and planetary cores \citep{youdin13}, makes cooling times too slow for VSI growth.  
 In this case cooling is too slow not just in the outer disk, but into  $< 1$ AU.

 An enhanced opacity makes optically thin cooling faster and radiative diffusion slower.  This shift favors VSI growth 
 in the outer disk at the expense of the inner disk, pushing the inner limit of VSI growth further out.
 The standard choice -- corresponding to Solar abundances in a standard MMSN disk -- allows the VSI 
 to grow over the widest range of relevant disk radii.

>>>>

\section{Summary and discussion}\label{summary}
In this paper, we have studied the vertical shear instability (VSI) with a focus the 
role of radiative cooling.  In turn we assess the viability of the VSI 
as an angular momentum transport mechanism in protoplanetary disks (PPDs).

@@

In applying our results to PPDs, we pay particular attention to the transition from 
optically thick radiative diffusion to optically thin Newtonian cooling.  The largest obstacle to VSI occurs in 
high density inner disk, $\lesssim 1$ -- 5AU, where radiative diffusion times are slow.  Shorter wavelength disturbances
 cool faster, so long as they remain optically thick.  In the inner disk, however, our VSI cooling criterion requires 
 wavelengths that are too short to drive significant transport.  The best hope for the VSI in inner disk appears to 
 be a low surface density, which speeds radiative diffusion by lengthening the photon mean free path.  This option naturally begs 
 the question of what accretion mechanism lowered the surface density in the first place.

 In the outer disk, the VSI tends to cool in the optically thin limit.  The main issue is whether the opacity 
 is high enough for optically thin cooling to be sufficiently fast.  Our standard opacity assumes a Solar abundances of small dust.
 In this case, the VSI can operate from $\sim 5$ -- 100AU.  
A factor of 10 reduction in the opacity, for instance by locking
 small dust into planetesimals and planetary cores \citep{youdin13}, makes cooling times too slow for VSI growth.  
 In this case cooling is too slow not just in the outer disk, but into  $< 1$ AU.

 An enhanced opacity makes optically thin cooling faster and radiative diffusion slower.  This shift favors VSI growth 
 in the outer disk at the expense of the inner disk, pushing the inner limit of VSI growth further out.
 The standard choice --- corresponding to Solar abundances in a standard MMSN disk --- allows the VSI 
 to grow over the widest range of relevant disk radii.
